---
title: "Missing Data Mechanism and Mean Imputation"
author: "Anthony Kamau"
output: pdf_document
---

# Income Dataset 

## Load the required libraries
```{r, message=FALSE}
library(MASS)
library(tidyverse)
library(ggplot2)

# set seed for reproducability
set.seed(45)
```


## Explore the data
```{r}
# Load data
income <- readRDS("income.rds")

# first 5 rows
head(income, 5)

# columns
names(income)

# five number summary
summary(income)

# Dataset consists of 3 variable i.e gender, age and income
# Average income and age is 2306 dollar and 43 years respectively
```


## Function for computing the probability fof a value being missing for each person 
using logistic regression model. This function takes a person's age, gender, 
and income, then weights each one by $\psi$ values, and returns a number between 0 
and 1 representing that person's chance of having a missing income value. 
Setting different $\psi$ values is exactly how the practical simulate:

    - NDD (all$\psi$ = 0, constant 0.5 for everyone)
    - SDD (nonzero psi1/psi2, zero psi3, depends only on observed age/gender)
    - UDD (nonzero psi3, depends on income itself).

```{r}
prob_mis <- function(psi0, psi1, psi2, psi3, age, gender, income) {
exp(psi0 + psi1 * age + psi2 * (gender == "Male") + psi3 * income) /
(1 + exp(psi0 + psi1 * age + psi2 * (gender == "Male") + psi3 * income))
}

```


# Not Data Dependent (NDD)
```{r}
income <-
  income |>
  mutate(misprob_ndd = prob_mis(psi0 = 0,
                                psi1 = 0,
                                psi2 = 0,
                                psi3 = 0,
                                age = age,
                                gender = gender,
                                income = income))

```

```{r}
income <- 
  income |>
  mutate(R_ndd = 1 - rbinom(n(), 1, misprob_ndd))
```


```{r}
income <- 
  income |>
  mutate(income_ndd = ifelse(R_ndd == 0, NA, income))
```


```{r}
table(is.na(income$income_ndd))
```


# Seen Data Dependent (SDD)
```{r}
income <-
  income |>
  mutate(misprob_sdd = prob_mis(
    psi0   = -3,
    psi1   = 0.05,
    psi2   = 0.5,
    psi3   = 0,
    age    = age,
    gender = gender,
    income = income))
```


```{r}
income |> 
  ggplot(aes(x = age, y = misprob_sdd, colour = gender)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    x = "Age", 
    y = "Missingness probability", 
    colour = "Gender",
    title = "Seen data dependent missingness"
  )

# missingness probability rises steadily as age increases, with males consistently 
# showing a higher probability than females at every age. 
# This is as expect given the parameters we set (psi1 = 0.05 for age, psi2 = 0.5 for male). 
# It confirms the SDD mechanism is working correctly because missingness depends
# on observed variables (age and gender).
# An 18-year-old female has roughly a 10% chance of missing income, while a 
# 70-year-old male has over 70%.
```

```{r}
income <- 
  income |>
  mutate(
    R_sdd = 1 - rbinom(n(), 1, misprob_sdd),
    income_sdd = ifelse(R_sdd == 0, NA, income)
  )
```


# Unseen Data Dependent (UDD)
```{r}
income <-
  income |>
  mutate(misprob_udd = prob_mis(
    psi0   = -7,
    psi1   = 0,
    psi2   = 0,
    psi3   = 0.003,
    age    = age,
    gender = gender,
    income = income
  ))
```


```{r}
income |> 
  ggplot(aes(x = income, y = misprob_udd)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(
    x = "Income", 
    y = "Missingness probability",
    title = "Unseen data dependent missingness"
  )

# Missingness probability climbs steeply as income itself increases, from about 10%
# at the lowest incomes to nearly 90% at the highest. 
# This is the defining feature of UDD where the probability of a value being missing
# depends on the value itself.
# High earners are much more likely to have their income hidden. 
```


```{r}
income <- 
  income |>
  mutate(
    R_udd = 1 - rbinom(n(), 1, misprob_udd),
    income_udd = ifelse(R_udd == 0, NA, income)
  )
```

# Visualize missing data

Visualization can help in understand the missing data mechanism by providng 
insight into the amount and pattern of missingness.
```{r, warning=FALSE, message=FALSE}
income |> 
  pivot_longer(
    cols = c(R_ndd, R_sdd, R_udd),
    names_to = "mechanism",
    values_to = "observed",
    names_prefix = "R_"
  ) |>
  mutate(observed = factor(ifelse(observed == 0, "missing", "observed"))) |> 
  ggplot(aes(x = str_to_upper(mechanism), y = income, fill = observed)) +
  geom_boxplot() + 
  theme_minimal() + 
  scale_fill_manual(values = c("seagreen", "lightblue")) +
  labs(
    x = "Missingness mechanism", 
    y = "Income value",
    colour = ""
  )

# These boxplots display the distributions of $Y_1$ for the observed and the 
# missing group under the three different missing data mechanisms. We see that 
# under NDD, the distributions are approximately similar, while they become 
# progressively more distinct under SDD and UDD.
```

# Mean Imputation
Why mean imputation, in general, is a bad idea

```{r}
# with across() we can apply a mutate over several columns

income <- 
  income |>
  mutate(across(
    .cols  = c(income_ndd, income_sdd, income_udd),
    .fns   = function(x) replace_na(x, mean(x, na.rm = TRUE)),
    .names = "imp_{.col}"
  ))
```


```{r}
income |> 
  pivot_longer(
    cols = c(income, starts_with("imp_")), 
    names_to = "mechanism", 
    values_to = "value"
  ) |> 
  group_by(mechanism) |> 
  summarize(mean = mean(value), sd = sd(value))
```


```{r, message=FALSE}
income |> 
  pivot_longer(
    cols = c(income, starts_with("imp_")), 
    names_to = "mechanism", 
    values_to = "value"
  ) |> 
  ggplot(aes(x = value, fill = mechanism)) +
  geom_histogram() +
  facet_wrap(vars(mechanism)) +
  theme_minimal() +
  scale_fill_viridis_d()

# The mean imputation creates an enormous "peak" in the middle that is not there
# in the real data.
# mean imputation shrinks the standard deviation and distorts the shape of the distribution,
# misrepresenting the real spread of incomes.
# this problem exists for all three mechanisms, which is why mean imputation is 
# described as "in general, a bad idea" regardless of what caused the missingness.
```

```{r}
head(income, 10)
```

